# The Normal Random Variable {#ChapNormal}

## Student Learning Objective

This chapter introduces a very important bell-shaped distribution known
as the Normal distribution. Computations associated with this
distribution are discussed, including the percentiles of the
distribution and the identification of intervals of subscribed
probability. The Normal distribution may serve as an approximation to
other distributions. We demonstrate this property by showing that under
appropriate conditions the Binomial distribution can be approximated by
the Normal distribution. This property of the Normal distribution will
be picked up in the next chapter where the mathematical theory that
establishes the Normal approximation is demonstrated. By the end of this
chapter, the student should be able to:

-   Recognize the Normal density and apply `R` functions for computing
    Normal probabilities and percentiles.

-   Associate the distribution of a Normal random variable with that of
    its standardized counterpart, which is obtained by centering and
    re-scaling.

-   Use the Normal distribution to approximate the Binomial
    distribution.

## The Normal Random Variable {#the-normal-random-variable}

The Normal distribution is the most important of all distributions that
are used in statistics. In many cases it serves as a generic model for
the distribution of a measurement. Moreover, even in cases where the
measurement is modeled by other distributions (i.e. Binomial, Poisson,
Uniform, Exponential, etc.) the Normal distribution emerges as an
approximation of the distribution of numerical characteristics of the
data produced by such measurements.

### The Normal Distribution

A Normal random variable has a continuous distribution over the sample
space of all numbers, negative or positive. We denote the Normal
distribution via “$X \sim \mathrm{Normal}(\mu, \sigma^2)$", where
$\mu = \Expec(X)$ is the expectation of the random variable and
$\sigma^2 = \Var(X)$ is it’s variance[^6_1].

Consider, for example, $X \sim \mathrm{Normal}(2,9)$. The density of the
distribution is presented in Figure \@ref(fig:Normal1). Observe that the
distribution is symmetric about the expectation 2. The random variable
is more likely to obtain its value in the vicinity of the expectation.
Values much larger or much smaller than the expectation are
substantially less likely.

```{r Normal1, fig.cap='The Normal(2,9) Distribution', echo=FALSE, message=FALSE,warning=FALSE, out.width = '60%', fig.align = "center"}
x.c = seq(-10,12,.1)
plot(c(0,x.c),c(0,dnorm(x.c,2,3)), type="l", ylab="Density",xlab="x")
x.c = seq(0,5,.1)
polygon(c(0,x.c,5),c(0,dnorm(x.c,2,3),0), density=10)
text(-5.2,.12,"P(0 < x < 5)")
library(shape)
Arrows(2,.06,-4, .11)

```

The density of the Normal distribution can be computed with the aid of
the function “`dnorm`". The cumulative probability can be computed with
the function “`pnorm`". For illustrating the use of the latter function,
assume that $X \sim \mathrm{Normal}(2,9)$. Say one is interested in the
computation of the probability $\Prob(0 < X \leq 5)$ that the random
variable obtains a value that belongs to the interval $(0,5]$. The
required probability is indicated by the marked area in
Figure \@ref(fig:Normal1). This area can be computed as the difference
between the probability $\Prob(X \leq 5)$, the area to the left of 5,
and the probability $\Prob(X \leq 0)$, the area to the left of 0:

```{r}
pnorm(5,2,3) - pnorm(0,2,3)
```

The difference is the indicated area that corresponds to the probability
of being inside the interval, which turns out to be approximately equal
to 0.589. Notice that the expectation $\mu$ of the Normal distribution
is entered as the second argument to the function. The third argument to
the function is the standard deviation, i.e. the square root of the
variance. In this example, the standard deviation is $\sqrt{9}=3$.

Figure \@ref(fig:Normal2) displays the densities of the Normal
distribution for the combinations $\mu= 0$, $\sigma^2 = 1$ (the *red*
line); $\mu = 2$, $\sigma^2 = 9$ (the *black* line); and $\mu = -3$,
$\sigma^2 = 1/4$ (the *green* line). Observe that the smaller the
variance the more concentrated is the distribution of the random
variable about the expectation.

```{r Normal2, fig.cap='The Normal Distribution for Various Values of $\\mu$ and $\\sigma^2$', echo=FALSE, message=FALSE,warning=FALSE, out.width = '60%', fig.align = "center"}
x.coord = seq(-10,10,.1)
plot(c(0,x.coord), c(0,dnorm(x.coord,0,1)), ylab ="Density", lty=1, xlab="x",ylim=c(0, .8), type="l")
lines(c(0,x.coord)-.03, c(0,dnorm(x.coord,2,3)), col=2)
lines(c(0,x.coord)+.03, c(0,dnorm(x.coord,-3,sqrt(.25))), col=3)
legend(4, .8, c("N(0,1)", "N(2,9)", "N(-3,0.25)"), col = c(1, 2, 3), lty = c(1, 1, 1))
```

```{example, exnormal1}
IQ tests are a popular (and controversial) mean for
measuring intelligence. They are produced as (weighted) average of a
response to a long list of questions, designed to test different
abilities. The score of the test across the entire population is set to
be equal to 100 and the standard deviation is set to 15. The
distribution of the score is Normal. Hence, if $X$ is the IQ score of a
random subject then $X \sim \mathrm{Normal}(100,15^2)$.
```

```{example, exnormal2}
Any measurement that is produced as a result of the
combination of many independent influencing factors is likely to poses
the Normal distribution. For example, the hight of a person is
influenced both by genetics and by the environment in which that person
grew up. Both the genetic and the environmental influences are a
combination of many factors. Thereby, it should not come as a surprise
that the heights of people in a population tend to follow the Normal
distribution.
```

### The Standard Normal Distribution

The standard normal distribution is a normal distribution of
standardized values, which are called $z$-scores. A $z$-score is the
original measurement measured in units of the standard deviation from
the expectation. For example, if the expectation of a Normal
distribution is 2 and the standard deviation is $3 = \sqrt{9}$, then the
value of 0 is 2/3 standard deviations smaller than (or to the left of)
the expectation. Hence, the $z$-score of the value 0 is -2/3. The
calculation of the $z$-score emerges from the equation:

$$(0 =)\; x = \mu + z \cdot \sigma\; (= 2 + z \cdot 3)$$ The $z$-score
is obtained by solving the equation

$$0 = 2 + z \cdot 3 \quad \Longrightarrow \quad z = (0-2)/3 = -2/3\;.$$
In a similar way, the $z$-score of the value $x = 5$ is equal to 1,
following the solution of the equation $5 = 2 + z\cdot 3$, which leads
to $z = (5-2)/3 = 1$.

The standard Normal distribution is the distribution of a standardized
Normal measurement. The expectation for the standard Normal distribution
is 0 and the variance is 1. When $X \sim N(\mu,\sigma^2)$ has a Normal
distribution with expectation $\mu$ and variance $\sigma^2$ then the
transformed random variable $Z = (X-\mu)/\sigma$ produces the standard
Normal distribution $Z\sim N(0,1)$. The transformation corresponds to
the reexpression of the original measurement in terms of a new “zero"
and a new unit of measurement. The new “zero" is the expectation of the
original measurement and the new unit is the standard deviation of the
original measurement.

Computation of probabilities associated with a Normal random variable
$X$ can be carried out with the aid of the standard Normal distribution.
For example, consider the computation of the probability
$\Prob(0 < X \leq 5)$ for $X \sim N(2, 9)$, that has expectation $\mu=2$
and standard deviation $\sigma = 3$. Consider $X$’s standardized values:
$Z = (X-2)/3$. The boundaries of the interval $[0,5]$, namely $0$ and
$5$, have standardized $z$-scores of $(0-2)/3=-2/3$ and $(5-2)/3 =1$,
respectively. Clearly, the original measurement $X$ falls between the
original boundaries ($0 < X \leq 5$) if, and only if, the standardized
measurement $Z$ falls between the standardized boundaries
($-2/3 < Z \leq 1$). Therefore, the probability that $X$ obtains a value
in the range $[0,5]$ is equal to the probability that $Z$ obtains a
value in the range $[-2/3,1]$.

```{r Normal3, fig.cap='The Standard Normal Distribution', echo=FALSE, message=FALSE, warning=FALSE, out.width = '60%', fig.align = "center"}

x.c = seq(-4,4,.1)
plot(x.c,dnorm(x.c,0,1), type="l", ylab="Density",xlab="x")
x.c = seq(-2/3,1,.1)
polygon(c(-2/3,x.c,1),c(0,dnorm(x.c,0,1),0), density=10)
text(-2.5,.35,"P((0-2)/3 < Z < (5-2)/3)")
Arrows(0,.15,-2, .3)

```

The function “`pnorm`" was used in the previous subsection in order to
compute that probability that $X$ obtains values between 0 and 5. The
computation produced the probability 0.5888522. We can repeat the
computation by the application of the same function to the standardized
values:

```{r}
pnorm((5-2)/3) - pnorm((0-2)/3)
```

The value that is being computed, the area under the graph for the
standard Normal distribution, is presented in Figure \@ref(fig:Normal3).
Recall that 3 arguments where specified in the previous application of
the function “`pnorm`": the $x$ value, the expectation, and the standard
deviation. In the given application we did not specify the last two
arguments, only the first one. (Notice that the output of the expression
“`(5-2)/3`" is a single number and, likewise, the output of the
expression “`(0-2)/3`" is also a single number.)

Most `R` function have many arguments that enables flexible application
in a wide range of settings. For convenience, however, default values
are set to most of these arguments. These default values are used unless
an alternative value for the argument is set when the function is
called. The default value of the second argument of the function
“`pnorm`" that specifies the expectation is “`mean=0`", and the default
value of the third argument that specifies the standard deviation is
“`sd=1`". Therefore, if no other value is set for these arguments the
function computes the cumulative distribution function of the standard
Normal distribution.

### Computing Percentiles

Consider the issue of determining the range that contains 95% of the
probability for a Normal random variable. We start with the standard
Normal distribution. Consult Figure \@ref(fig:Normal4). The figure
displays the standard Normal distribution with the central region
shaded. The area of the shaded region is 0.95.

We may find the $z$-values of the boundaries of the region, denoted in
the figure as $z_0$ and $z_1$ by the investigation of the cumulative
distribution function. Indeed, in order to have 95% of the distribution
in the central region one should leave out 2.5% of the distribution in
each of the two tails. That is, 0.025 should be the area of the unshaded
region to the right of $z_1$ and, likewise, 0.025 should be the area of
the unshaded region to the left of $z_0$. In other words, the cumulative
probability up to $z_0$ should be 0.025 and the cumulative distribution
up to $z_1$ should be 0.975.

In general, given a random variable $X$ and given a percent $p$, the $x$
value with the property that the cumulative distribution up to $x$ is
equal to the probability $p$ is called the $p$-*percentile* of the
distribution. Here we seek the 2.5%-percentile and the 97.5%-percentile
of the standard Normal distribution.

```{r Normal4, fig.cap='Central 95$\\%$ of the Standard Normal Distribution',echo=FALSE, message=FALSE, warning=FALSE, out.width = '60%', fig.align = "center"}
v1 <- c(-4,-1.96,0,1.96,4)
z0 = expression(paste("z" [0], " = -1.96"))
z1 = expression(paste("z" [1], " = 1.96"))
v2 <- c("-4",z0,"0",z1,"4")
x.c = seq(-4,4,.1)
plot(x.c,dnorm(x.c,0,1), type="l", ylab="Density",xlab="z",xaxt = "n")
axis(side = 1, at = v1, labels = v2, tck=-.05)
x.c = seq(-1.96,1.96,.1)
polygon(c(-1.96,x.c,1.96),c(0,dnorm(x.c,0,1),0), density=10)
text(-2.5,.35,expression(paste("P(z" [0], "<Z<z" [1], ") = 0.95")), cex=.8)
Arrows(0,.15,-2, .3)

```

The percentiles of the Normal distribution are computed by the function
“`qnorm`". The first argument to the function is a probability (or a
sequence of probabilities), the second and third arguments are the
expectation and the standard deviations of the normal distribution. The
default values to these arguments are set to 0 and 1, respectively.
Hence if these arguments are not provided the function computes the
percentiles of the standard Normal distribution. Let us apply the
function in order to compute $z_1$ and $z_0$:

```{r}
qnorm(0.975)
qnorm(0.025)
```

Observe that $z_1$ is practically equal to 1.96 and
$z_0 = -1.96 = -z_1$. The fact that $z_0$ is the negative of $z_1$
results from the symmetry of the standard Normal distribution about 0.
As a conclusion we get that for the standard Normal distribution 95% of
the probability is concentrated in the range $[-1.96, 1.96]$.

```{r Normal5, fig.cap='Central 95$\\%$ of the Normal(2,9) Distribution', echo=FALSE, message=FALSE, warning=FALSE, out.width = '60%', fig.align = "center"}
v1 <- c(-10,-7,-3.9,0,5,7.9,10)
z0 = expression(paste("x" [0], " = ?"))
z1 = expression(paste("x" [1], " = ?"))
v2 <- c("-10","-7",z0,"0","5",z1,"10")
x.c = seq(-10,12,.1)
plot(x.c,dnorm(x.c,2,3), type="l", ylab="Density",xlab="x",xaxt = "n")
axis(side = 1, at = v1, labels = v2, tck=-.05)
x.c = seq(-3.9,7.9,.1)
polygon(c(-3.9,x.c,7.9),c(0,dnorm(x.c,2,3),0), density=10)
text(-6,.1,expression(paste("P(x" [0], "<X<x" [1], ") = 0.95")), cex=.8)
Arrows(2,.03,-5, .09)
```

The problem of determining the central range that contains 95% of the
distribution can be addresses in the context of the original measurement
$X$ (See Figure \@ref(fig:Normal5)). We seek in this case an interval
centered at the expectation 2, which is the center of the distribution
of $X$, unlike 0 which was the center of the standardized values $Z$.
One way of solving the problem is via the application of the function
“`qnorm`" with the appropriate values for the expectation and the
standard deviation:

```{r}
qnorm(0.975,2,3)
qnorm(0.025,2,3)
```

Hence, we get that $x_0 = -3.88$ has the property that the total
probability to its left is 0.025 and $x_1 = 7.88$ has the property that
the total probability to its right is 0.025. The total probability in
the range $[-3.88, 7.88]$ is 0.95.

An alternative approach for obtaining the given interval exploits the
interval that was obtained for the standardized values. An interval
$[-1.96,1.96]$ of standardized $z$-values corresponds to an interval
$[2 - 1.96 \cdot 3, 2 + 1.96\cdot 3]$ of the original $x$-values:

```{r}
2 + qnorm(0.975)*3
2 + qnorm(0.025)*3
```

Hence, we again produce the interval $[-3.88,7.88]$, the interval that
was obtained before as the central interval that contains 95% of the
distribution of the $\mathrm{Normal}(2,9)$ random variable.

In general, if $X \sim N(\mu,\sigma)$ is a Normal random variable then
the interval $[\mu - 1.96 \cdot \sigma, \mu + 1.96 \cdot \sigma]$
contains 95% of the distribution of the random variable. Frequently one
uses the notation $\mu \pm 1.96 \cdot \sigma$ to describe such an
interval.

### Outliers and the Normal Distribution

Consider, next, the computation of the interquartile range in the Normal
distribution. Recall that the interquartile range is the length of the
central interval that contains 50% of the distribution. This interval
starts at the first quartile (Q1), the value that splits the
distribution so that 25% of the distribution is to the left of the value
and 75% is to the right of it. The interval ends at the third quartile
(Q3) where 75% of the distribution is to the left and 25% is to the
right.

For the standard Normal the third and first quartiles can be computed
with the aid of the function “`qnorm`":

```{r}
qnorm(0.75)
qnorm(0.25)
```

Observe that for the standard Normal distribution one has that 75% of
the distribution is to the left of the value 0.6744898, which is the
third quartile of this distribution. Likewise, 25% of the standard
Normal distribution are to the left of the value -0.6744898, which is
the first quartile. the interquartile range is the length of the
interval between the third and the first quartiles. In the case of the
standard Normal distribution this length is equal to
$0.6744898 - (-0.6744898) = 1.348980$.

In Chapter \@ref(ChapDescriptiveStat) we considered box plots as a mean for
the graphical display of numerical data. The box plot includes a
vertical rectangle that initiates at the first quartile and ends at the
third quartile, with the median marked within the box. The rectangle
contains 50% of the data. Whiskers extends from the ends of this
rectangle to the smallest and to the largest data values that are not
outliers. Outliers are values that lie outside of the normal range of
the data. Outliers are identified as values that are more then 1.5 times
the interquartile range away from the ends of the central rectangle.
Hence, a value is an outlier if it is larger than the third quartile
plus 1.5 times the interquartile range or if it is less than the first
quartile minus 1.5 times the interquartile range.

How likely is it to obtain an outlier value when the measurement has the
standard Normal distribution? We obtained that the third quartile of the
standard Normal distribution is equal to 0.6744898 and the first
quartile is minus this value. The interquartile range is the difference
between the third and first quartiles. The upper and lower thresholds
for the defining outliers are:

```{r}
qnorm(0.75) + 1.5*(qnorm(0.75)-qnorm(0.25))
qnorm(0.25) - 1.5*(qnorm(0.75)-qnorm(0.25))
```

Hence, a value larger than 2.697959 or smaller than -2.697959 would be
identified as an outlier.

The probability of being less than the upper threshold 2.697959 in the
standard Normal distribution is computed with the expression
“`pnorm(2.697959)`". The probability of being above the threshold is 1
minus that probability, which is the outcome of the expression
“`1-pnorm(2.697959)`".

By the symmetry of the standard Normal distribution we get that the
probability of being below the lower threshold -2.697959 is equal to the
probability of being above the upper threshold. Consequently, the
probability of obtaining an outlier is equal to twice the probability of
being above the upper threshold:

```{r}
2*(1-pnorm(2.697959))
```

We get that for the standard Normal distribution the probability of an
outlier is approximately 0.7%.

## Approximation of the Binomial Distribution

The Normal distribution emerges frequently as an approximation of the
distribution of data characteristics. The probability theory that
mathematically establishes such approximation is called the Central
Limit Theorem and is the subject of the next chapter. In this section we
demonstrate the Normal approximation in the context of the Binomial
distribution.

### Approximate Binomial Probabilities and Percentiles

Consider, for example, the probability of obtaining between 1940 and
2060 heads when tossing 4,000 fair coins. Let $X$ be the total number of
heads. The tossing of a coin is a trial with two possible outcomes:
“Head" and “Tail." The probability of a “Head" is 0.5 and there are
4,000 trials. Let us call obtaining a “Head" in a trial a “Success".
Observe that the random variable $X$ counts the total number of
successes. Hence, $X \sim \mathrm{Binomial}(4000,0.5)$.

The probability $\Prob(1940 \leq X \leq 2060)$ can be computed as the
difference between the probability $\Prob(X \leq 2060)$ of being less or
equal to 2060 and the probability $\Prob(X < 1940)$ of being strictly
less than 1940. However, 1939 is the largest integer that is still
strictly less than the integer 1940. As a result we get that
$\Prob(X < 1940) = \Prob(X \leq 1939)$. Consequently,
$\Prob(1940 \leq X \leq 2060) = \Prob(X \leq 2060) - \Prob(X \leq 1939)$.

Applying the function “`pbinom`" for the computation of the Binomial
cumulative probability, namely the probability of being less or equal to
a given value, we get that the probability in the range between 1940 and
2060 is equal to

```{r}
pbinom(2060,4000,0.5) - pbinom(1939,4000,0.5)
```

This is an exact computation. The Normal approximation produces an
approximate evaluation, not an exact computation. The Normal
approximation replaces Binomial computations by computations carried out
for the Normal distribution. The computation of a probability for a
Binomial random variable is replaced by computation of probability for a
Normal random variable that has the same expectation and standard
deviation as the Binomial random variable.

Notice that if $X \sim \mathrm{Binomial}(4000,0.5)$ then the expectation
is $\Expec(X) = 4,000 \times 0.5 = 2,000$ and the variance is
$\Var(X) = 4,000 \times 0.5 \times 0.5 = 1,000$, with the standard
deviation being the square root of the variance. Repeating the same
computation that we conducted for the Binomial random variable, but this
time with the function “`pnorm`" that is used for the computation of the
Normal cumulative probability, we get:

```{r}
mu <- 4000*0.5
sig <- sqrt(4000*0.5*0.5)
pnorm(2060,mu,sig) - pnorm(1939,mu,sig)
```

Observe that in this example the Normal approximation of the probability
(0.9442441) agrees with the Binomial computation of the probability
(0.9442883) up to 3 significant digits.

Normal computations may also be applied in order to find approximate
percentiles of the Binomial distribution. For example, let us identify
the central region that contains for a $\mathrm{Binomial}(4000,0.5)$
random variable (approximately) 95% of the distribution. Towards that
end we can identify the boundaries of the region for the Normal
distribution with the same expectation and standard deviation as that of
the target Binomial distribution:

```{r}
qnorm(0.975,mu,sig)
qnorm(0.025,mu,sig)
```

After rounding to the nearest integer we get the interval $[1938,2062]$
as a proposed central region.

In order to validate the proposed region we may repeat the computation
under the actual Binomial distribution:

```{r}
qbinom(0.975,4000,0.5)
qbinom(0.025,4000,0.5)
```

Again, we get the interval $[1938,2062]$ as the central region, in
agreement with the one proposed by the Normal approximation. Notice that
the function “`qbinom`" produces the percentiles of the Binomial
distribution. It may not come as a surprise to learn that “`qpois`",
“`qunif`", “`qexp`" compute the percentiles of the Poisson, Uniform and
Exponential distributions, respectively.

The ability to approximate one distribution by the other, when
computation tools for both distributions are handy, seems to be of
questionable importance. Indeed, the significance of the Normal
approximation is not so much in its ability to approximate the Binomial
distribution as such. Rather, the important point is that the Normal
distribution may serve as an approximation to a wide class of
distributions, with the Binomial distribution being only one example.
Computations that are based on the Normal approximation will be valid
for all members in the class of distributions, including cases where we
don’t have the computational tools at our disposal or even in cases
where we do not know what the exact distribution of the member is! As
promised, a more detailed discussion of the Normal approximation in a
wider context will be presented in the next chapter.

On the other hand, one need not assume that any distribution is well
approximated by the Normal distribution. For example, the distribution
of wealth in the population tends to be skewed, with more than 50% of
the people possessing less than 50% of the wealth and small percentage
of the people possessing the majority of the wealth. The Normal
distribution is not a good model for such distribution. The Exponential
distribution, or distributions similar to it, may be more appropriate.

### Continuity Corrections

In order to complete this section let us look more carefully at the
Normal approximations of the Binomial distribution.

```{r Normal6, fig.cap='Normal Approximation of the Binomial Distribution', echo=FALSE, message=FALSE,warning=FALSE, out.width = '60%', fig.align = "center"}
x.c = 7:30
plot(x.c,dbinom(x.c,30,.3), type="h", ylab="Pr",xlab="x",xlim=c(0,30))
x.c = 1:6
lines(x.c,dbinom(x.c,30,.3), type="h", ylab="Pr",xlab="x",col=2)
x.c = seq(0,30,.1)
lines(x.c,dnorm(x.c,30*0.3,sqrt(30*0.3*0.7)))
x.c = seq(0,6.5,.1)
polygon(c(0,x.c,6.5),c(0,dnorm(x.c,30*0.3,sqrt(30*0.3*0.7)),0), density=10, col=2)

```

In principle, the Normal approximation is valid when $n$, the number of
independent trials in the Binomial distribution, is large. When $n$ is
relatively small the approximation may not be so good. Indeed, take
$X \sim \mathrm{Binomial}(30,0.3)$ and consider the probability
$\Prob(X \leq 6)$. Compare the actual probability to the Normal
approximation:

```{r}
pbinom(6,30,0.3)
pnorm(6,30*0.3,sqrt(30*0.3*0.7))
```

The Normal approximation, which is equal to 0.1159989, is not too close
to the actual probability, which is equal to 0.1595230.

A naïve application of the Normal approximation for the
$\mathrm{Binomial}(n,p)$ distribution may not be so good when the number
of trials $n$ is small. Yet, a small modification of the approximation
may produce much better results. In order to explain the modification
consult Figure \@ref(fig:Normal6) where you will find the bar plot of the
Binomial distribution with the density of the approximating Normal
distribution superimposed on top of it. The target probability is the
sum of heights of the bars that are painted in *red*. In the naïve
application of the Normal approximation we used the area under the
normal density which is to the left of the bar associated with the value
$x=6$.

Alternatively, you may associate with each bar located at $x$ the area
under the normal density over the interval $[x-0.5, x+0.5]$. The
resulting correction to the approximation will use the Normal
probability of the event $\{X \leq 6.5\}$, which is the area shaded in
*red*. The application of this approximation, which is called
*continuity correction* produces:

```{r}
pnorm(6.5,30*0.3,sqrt(30*0.3*0.7))
```

Observe that the corrected approximation is much closer to the target
probability, which is 0.1595230, and is substantially better that the
uncorrected approximation which was 0.1159989. Generally, it is
recommended to apply the continuity correction to the Normal
approximation of a discrete distribution.

Consider the $\mathrm{Binomial}(n,p)$ distribution. Another situation
where the Normal approximation may fail is when $p$, the probability of
“Success" in the Binomial distribution, is too close to 0 (or too close
to 1). Recall, that for large $n$ the Poisson distribution emerged as an
approximation of the Binomial distribution in such a setting. One may
expect that when $n$ is large and $p$ is small then the Poisson
distribution may produce a better approximation of a Binomial
probability. When the Poisson distribution is used for the approximation
we call it a *Poisson Approximation*.

Let us consider an example. Let us analyze 3 Binomial distributions. The
expectation in all the distributions is equal to 2 but the number of
trials, $n$, vary. In the first case $n=20$ (and hence $p=0.1$), in the
second $n=200$ (and $p=0.01$), and in the third $n=2,000$ (and
$p=0.001$. In all three cases we will be interested in the probability
of obtaining a value less or equal to 3.

The Poisson approximation replaces computations conducted under the
Binomial distribution with Poisson computations, with a Poisson
distribution that has the same expectation as the Binomial. Since in all
three cases the expectation is equal to 2 we get that the same Poisson
approximation is used to the three probabilities:

```{r}
ppois(3,2)
```

The actual Binomial probability in the first case ($n=20$, $p=0.1$) and
a Normal approximation thereof are:

```{r}
pbinom(3,20,0.1)
pnorm(3.5,2,sqrt(20*0.1*0.9))
```

Observe that the Normal approximation (with a continuity correction) is
better than the Poisson approximation in this case.

In the second case ($n=200$, $p=0.01$) the actual Binomial probability
and the Normal approximation of the probability are:

```{r}
pbinom(3,200,0.01)
pnorm(3.5,2,sqrt(200*0.01*0.99))
```

Observe that the Poisson approximation that produces 0.8571235 is
slightly closer to the target than the Normal approximation. The greater
accuracy of the Poisson approximation for the case where $n$ is large
and $p$ is small is more pronounced in the final case ($n=2000$,
$p=0.001$) where the target probability and the Normal approximation
are:

```{r}
pbinom(3,2000,0.001)
pnorm(3.5,2,sqrt(2000*0.001*0.999))
```

Compare the actual Binomial probability, which is equal to 0.8572138, to
the Poisson approximation that produced 0.8571235. The Normal
approximation, 0.8556984, is slightly off, but is still acceptable.

## Exercises {#Normal4}

```{exercise}
Consider the problem of establishing regulations
concerning the maximum number of people who can occupy a lift. In
particular, we would like to assess the probability of exceeding maximal
weight when 8 people are allowed to use the lift simultaneously and
compare that to the probability of allowing 9 people into the lift.

Assume that the total weight of 8 people chosen at random follows a
normal distribution with a mean of 560kg and a standard deviation of
57kg. Assume that the total weight of 9 people chosen at random follows
a normal distribution with a mean of 630kg and a standard deviation of
61kg.

1.  What is the probability that the total weight of 8 people exceeds
    650kg?

2.  What is the probability that the total weight of 9 people exceeds
    650kg?

3.  What is the central region that contains 80% of distribution of the
    total weight of 8 people?

4.  What is the central region that contains 80% of distribution of the
    total weight of 9 people?

```


```{exercise}
Assume $X \sim \mbox{Binomial}(27,0.32)$. We are
interested in the probability $\Prob(X > 11)$.

1.  Compute the (exact) value of this probability.

2.  Compute a Normal approximation to this probability, without a
    continuity correction.

3.  Compute a Normal approximation to this probability, with a
    continuity correction.

4.  Compute a Poisson approximation to this probability.

```

## Summary

### Glossary {#glossary .unnumbered}

Normal Random Variable:

:   A bell-shaped distribution that is frequently used to model a
    measurement. The distribution is marked with
    $\mathrm{Normal}(\mu,\sigma^2)$.

Standard Normal Distribution:

:   The $\mathrm{Normal}(0,1)$. The distribution of standardized Normal
    measurement.

Percentile:

:   Given a percent $p \cdot 100\%$ (or a probability $p$), the value
    $x$ is the percentile of a random variable $X$ if it satisfies the
    equation $\Prob(X \leq x) = p$.

Normal Approximation of the Binomial:

:   Approximate computations associated with the Binomial distribution
    with parallel computations that use the Normal distribution with the
    same expectation and standard deviation as the Binomial.

Poisson Approximation of the Binomial:

:   Approximate computations associated with the Binomial distribution
    with parallel computations that use the Poisson distribution with
    the same expectation as the Binomial.

### Discuss in the Forum {#discuss-in-the-forum .unnumbered}

Mathematical models are used as tools to describe reality. These models
are supposed to characterize the important features of the analyzed
phenomena and provide insight. Random variables are mathematical models
of measurements. Some people claim that there should be a perfect match
between the mathematical characteristics of a random variable and the
properties of the measurement it models. Other claim that a partial
match is sufficient. What is your opinion?

When forming your answer to this question you may give an example of a
situation from you own field of interest for which a random variable can
serve as a model. Identify discrepancies between the theoretical model
and actual properties of the measurement. Discuss the appropriateness of
using the model in light of these discrepancies.

Consider, for example, testing IQ. The score of many IQ tests are
modeled as having a Normal distribution with an expectation of 100 and a
standard deviation of 15. The sample space of the Normal distribution is
the entire line of real numbers, including the negative numbers. In
reality, IQ tests produce only positive values.

[^6_1]: If $X \sim \mbox{Normal}(\mu,\sigma^2)$ then the density of $X$ is
    given by the formula
    $f(x) = \exp\{-\frac{(x-\mu)^2}{2 \sigma^2}\}/\sqrt{2 \pi \sigma^2}$,
    for all $x$.
